Posterior Retrodictive Checks

### 2024-11-15 paper/paper-papaja-revising.Rmd
resubmission

affiliation:
  - id            : "1"
    institution   : "Arizona State University"

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "7531"

@@

Similarly, Unkelbach and Rom [-@unkelbach.rom2017] examined both implied and contradictory statements. They found that exposure to implied statements increased truth ratings (Exp 4) and that statements saw a significant decrease in truth ratings following incidental exposure to their contradictions (Exp. 3). Crucially, the results of this experiment appear to preclude any purely fluency-based account of reduced truth ratings following incidental exposure to contradictory statements: These researchers found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus, even if an appropriate notion of "conceptual disfluency" could be developed, it is unlikely it could be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of incidental exposure [@garcia-marques.etal2015; @gilbert1991; @silva.etal2017; @unkelbach.rom2017].

However, these last two sets of findings must be qualified by the exposure context used in these studies: These researchers asked participants to make incidental judgments about these statements but did not warn them that the statements could be true or false [@silva.etal2017; @unkelbach.rom2017]. Thus, the pragmatics of the task are unclear, and participants might have believed these statements because they understood them to be presented as facts.[^urnote]

[^urnote]: Unkelbach and Rom [-@unkelbach.rom2017] did test a "warning" condition in another experiment of a similar design (Exp. 1), but here did not find a significant reduction in truth judgments for contradictions of previously-seen statements.

@@

Existing research examining the effects of incidental exposure on implied statements has two key limitations. First, it has often been conducted in ambiguous contexts where participants might not understand that statements might be true or false, and that therefore do not always clearly induce incidental exposure. We address this through study designs inducing clear cases of incidental exposure. Second, prior work has only examined effects on direct contradictions of originally-seen statements (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of incidental exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether incidental exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

>>>>

affiliation:
  - id            : "1"
    institution   : "University of California, Irvine"
  - id            : "2"
    institution   : "Arizona State University"

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "7658"

@@

Similarly, Unkelbach and Rom [-@unkelbach.rom2017] examined both implied and contradictory statements. Consistent with the hypothesized Illusory Implication effect, they found that exposure to statements increased truth ratings for novel implied statements (Exp. 4) and that statements saw a significant decrease in truth ratings following incidental exposure to their contradictions (Exp. 3). Crucially, the results of this experiment appear to preclude any purely fluency-based account of reduced truth ratings following exposure to contradictory statements: These researchers found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus, even if an appropriate notion of "conceptual disfluency" could be developed, it is unlikely it could be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of incidental exposure [@garcia-marques.etal2015; @gilbert1991; @silva.etal2017; @unkelbach.rom2017].

However, these last two sets of findings must be qualified by the exposure context used in these studies: These researchers exposed participants to statements by asking them to make innocuous judgments about these statements (e.g., "how interesting is this statement?") but did not warn them prior to this exposure that the statements could be true or false [@silva.etal2017; @unkelbach.rom2017]. The pragmatics of this task are therefore ambiguous: participants might have imagined they were being asked to rate the interesting-ness of a set of entirely true statements, and might have then believed those statements for that reason. In the absence of strong and clear task pragmatics (e.g. a true or false quiz), explicit warnings are key to establishing incidental exposure as we have defined it here.

@@

Existing research examining the effects of incidental exposure on implied statements has two key limitations. First, it has often been conducted in ambiguous contexts where participants might not understand that statements might be true or false, and that therefore do not always clearly induce incidental exposure, as we have defined it here. We address this in our study designs, by paying careful attention to the pragmatics of the exposure tasks to ensure they constitute clear cases of incidental exposure. Second, prior work has primarily examined effects on direct contradictions of originally-seen statements (or paraphrases of direct contradictions) [but see @unkelbach.rom2017]. Here, we sought to further test for an "illusory implication effect" by testing the effect of incidental exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether incidental exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).
